A Simple Type Theory for Category Theory

### 2022-04-28 paper/paper.tex
The interesting part of the completeness proof

    \inferrule{\Gamma \vdash \cat C : \isaCat}{\Gamma \vdash \alpha: \cat C \boundary}

    \inferrule{\Gamma \vdash \cat C : \isaCat\and
      \Gamma\vdash \cat D : \isaCat}{\Gamma \vdash \alpha: \cat C;\beta:\cat D \boundary}\\

    \inferrule*
        {\Gamma \pipe \alpha':\cat C' \vdash A : \cat C}
        {\Gamma \vdash a/\alpha : \alpha':\cat C' \to \alpha:\cat C}

    \inferrule*
    {\Gamma \pipe d^-\Xi \vdash a : \cat C\and
     \Gamma \pipe d^+\Xi \vdash b : \cat D
    }
    {\Gamma \vdash (a/\alpha;b/\beta) : \Xi \to (\alpha:\cat C;\beta:\cat D)}\\

@@

    \inferrule
    {\Gamma \vdash \Phi \isavectx\and
    \Gamma \vdash \Psi \isavectx \and
    \Gamma \pipe d^+\Phi; d^- \Psi \vdash R \isaSet}
    {\Gamma \vdash \Phi, x:R, \Psi \isavectx}

@@

    \inferrule
    {\Psi = \Psi_1 \jnctx \Psi_M \jnctx \Psi_2\and
      \Gamma \vdash \phi_1 : \Psi_1 \to \Phi_1\and
      \Gamma \pipe \Psi_s \vdash s : R[d^+\phi_1;d^-\phi_2] \and
      \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
      }
    {\Gamma \vdash \phi_1,t/x,\phi_2 : \Psi \to \Phi_1,x:R,\Phi_2}
  \end{mathpar}
  \begin{align*}
    \underline{\alpha:C} &= \alpha:C\\
    \underline{\Phi,x:R,\Psi} &= d^-\Phi; d^+{\Psi}
  \end{align*}
  \begin{align*}
    d^{\pm}(\alpha:\cat C) &= \alpha : \cat C\\
    d^{-}(\Phi,x:R,\Psi) &= d^-\Phi\\
    d^{+}(\Phi,x:R,\Psi) &= d^-\Psi
  \end{align*}
    \begin{align*}
    d^{\pm}(a/\alpha) &= a/\alpha\\
    d^{-}(\phi,t/x,\psi) &= d^-\phi\\
    d^{+}(\phi,t/x,\psi) &= d^+\psi\\
  \end{align*}
  \caption{Basic Rules of Judgment Theory}
  \label{fig:contexts}
\end{figure}

@@

%% \appendix

%% TODO: complete substitution
%% \begin{figure}
%%   \begin{mathpar}
%%       (\homrallXYtoZ \alpha R P)[\xi] = \homrallXYtoZ \alpha {R[d^+\xi;\alpha/\alpha]} {P[\alpha/\alpha;d^-\xi]}

@@

%%   \end{mathpar}
%%   \caption{Substitution}

%% \end{figure}

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule
%%     {}
%%     {\Gamma\pipe \cdot \vdash \Sort \isaSet}

%%     \inferrule
%%     {x:S \in \Gamma}
%%     {\Gamma \pipe \cdot \vdash x : S}

%%     \inferrule
%%     {\Gamma\pipe\cdot \vdash C : \Sort}
%%     {\Gamma \pipe \alpha:\cat C \vdash \alpha : \cat C}

%%     \inferrule
%%     {\Gamma \pipe \alpha:C,\beta:D \vdash R \isaSet}
%%     {\Gamma\pipe \alpha:C, x:R, \beta: D \vdash x : R}

%%     \inferrule{\forall x:A \in \Gamma. \Delta \vdash \gamma(x) : A}{\gamma : \Delta \to \Gamma}\\

%%     \inferrule*%% [right=NullaryBoundaryMap]
%%                {\Gamma \isadtctx}{\Gamma \vdash \cdot : \cdot \to \cdot}\and
%%     \inferrule*{\Gamma \pipe \alpha':C' \vdash A : C}{\Gamma \vdash A/\alpha : \alpha':C' \to \alpha:C}

%%     \inferrule*
%%     {\Gamma \pipe \alpha':C' \vdash A : C\and
%%       \Gamma \pipe \beta':D' \vdash B : D
%%     }
%%     {\Gamma \vdash (A/\alpha;B/\beta) : (\alpha':C';\beta':D') \to (\alpha:C;\beta':D')}\\

%%     \inferrule*%% [right=NullCtxMap]
%%                {\Gamma \isadtctx}{\Gamma \vdash \cdot : \cdot \to \cdot}

%%     \inferrule*{\Gamma \pipe \alpha':C' \vdash A : C}{\Gamma \vdash A/\alpha : \alpha':C' \to \alpha:C}

%%     \inferrule
%%     {\Psi = \Psi_1 \jnctx \Psi_M \jnctx \Psi_2\and
%%       \Gamma \vdash \phi_1 : \Psi_1 \to \Phi_1\and
%%       \Gamma \pipe \Psi_M \vdash M : R[d^+\phi_1;d^-\phi_2] \and
%%       \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
%%       }
%%     {\Gamma \vdash \phi_1,M/x,\phi_2 : \Psi \to \Phi_1,x:R,\Phi_2}
%%   \end{mathpar}
%%   \caption{Substitutions in SCT}
%%   \label{fig:basic-rules}
%% \end{figure}

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule
%%     {}
%%     {\cdot : \cdot \to \cdot}

%%     \inferrule
%%     {\gamma : \Delta \to \Gamma \and \Delta \vdash M : A[\gamma]}
%%     {\gamma,M/X : \Delta \to \Gamma,X:A}

%%     \inferrule
%%     {\Gamma \vdash \Phi \isavectx \and
%%     \Gamma \vdash \Psi \isavectx \and
%%     d^+\Phi = d^- \Psi}
%%     {\Gamma \vdash \Phi \jnctx \Psi \isavectx}

%%     \inferrule
%%     {\Gamma \vdash \phi : \Phi' \to \Phi \and
%%      \Gamma \vdash \psi : \Psi' \to \Psi\and
%%      d^+\phi = d^- \psi}
%%     {\Gamma \vdash \phi \jnctx \psi : \Phi' \jnctx \Psi' \to \Phi \jnctx \Psi}

%%     \inferrule*[right=ObSubst]
%%     {\alpha : \cat C \vdash B : \cat D\and
%%     \beta : \cat D \vdash A : \cat E}
%%     {\alpha : \cat C \vdash A[B/\beta] : \cat E}\\

%%     \inferrule*[right=Everything Subst]{\gamma : \Delta \to \Gamma \and \Gamma J}{\Delta J}

%%     \inferrule{\Gamma \vdash \xi : Z \to \Xi\and \Gamma \pipe \Xi \vdash R \isaSet}{\Gamma \pipe Z \vdash R[\xi] \isaSet}

%%     \inferrule{\Gamma \vdash \phi : \Psi \to \Phi\and \Gamma \pipe \Phi \vdash M : R}{\Gamma \pipe \Psi \vdash M[\phi] : R[\underline \phi]}
%%   \end{mathpar}
%%   \begin{align*}
%%     (\Phi \jnctx \Psi) \jnctx \Sigma &= \Phi \jnctx (\Psi \jnctx \Sigma)\\
%%     \Phi \jnctx \alpha:C &= \Phi\\
%%     \alpha:C \jnctx \Psi &= \Psi\\\\
%%     d^\pm(A/\alpha) &= A/\alpha\\
%%     d^+(\phi_1,M/x,\phi_2) &= d^+ \phi_2\\
%%     d^-(\phi_1,M/x,\phi_2) &= d^- \phi_1\\\\
%%     \underline \cdot &= \cdot\\
%%     \underline {A/\alpha} &= A/\alpha\\
%%     \underline {\phi_1,M/x,\phi_2} &= d^-\phi_1;d^+\phi_2
%%   \end{align*}
%%   \caption{Admissible Rules}
%% \end{figure}

>>>>

@@

    \inferrule
    {\Gamma \vdash \Phi \isavectx\and
    \Gamma \pipe d^+\Phi; \beta:\cat D \vdash R \isaSet}
    {\Gamma \vdash \Phi, x:R, \beta:\cat D \isavectx}

@@

    % horizontal identity
    \inferrule
    {\Gamma\pipe \beta:\cat C \vdash a : \cat C}
    {\Gamma \pipe \beta:\cat C \vdash a/\alpha : (\alpha : \cat C)}

    \inferrule
    {\Psi = \Psi_1 \jnctx \Psi_2\and
      \Gamma\pipe\Psi_1 \vdash \phi : \Phi\and
      \Gamma \pipe \Psi_2 \vdash t : R[d^+\phi;b/\beta] \and
      \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
      }
    {\Gamma\pipe\Psi \vdash \phi,t/x,b/\beta :: (\Phi,x:R,\beta:\cat D)}
  \end{mathpar}
  \begin{align*}
    \underline{\alpha:C} &= \alpha:C\\
    \underline{\Phi,x:R,\Psi} &= d^-\Phi; d^+{\Psi}
  \end{align*}
  \begin{align*}
    d^{\pm}(\alpha:\cat C) &= \alpha : \cat C\\
    d^{-}(\Phi,x:R,\Psi) &= d^-\Phi\\
    d^{+}(\Phi,x:R,\Psi) &= d^-\Psi
  \end{align*}
  \begin{align*}
    d^{\pm}(a/\alpha) &= a/\alpha\\
    d^{-}(\phi,t/x,\psi) &= d^-\phi\\
    d^{+}(\phi,t/x,\psi) &= d^+\psi\\
  \end{align*}
  \caption{Basic Rules of Judgment Theory}
  \label{fig:contexts}
\end{figure}

@@

\appendix

\section{Completeness}

%% \begin{figure}
%%   \begin{mathpar}
%%       (\homrallXYtoZ \alpha R P)[\xi] = \homrallXYtoZ \alpha {R[d^+\xi;\alpha/\alpha]} {P[\alpha/\alpha;d^-\xi]}

@@

%%   \end{mathpar}
%%   \caption{Substitution}
%% \end{figure
%% }

\begin{definition}
  We define composition of transformation contexts $\Phi \jnctx \Psi$ by induction on $\Psi$
  \begin{align*}
    \Phi \jnctx \beta:\cat D &= \Phi    \\
    \Phi \jnctx \Psi,x:R,\beta:\cat D &= (\Phi\jnctx\Psi),x:R,\beta:\cat D
  \end{align*}
\end{definition}

It follows by an inductive argument that this definition is
associative and unital.

\begin{definition}
  We define the horizontal composition of transformation substitutions
  $\phi \jnctx \psi$ whenever $d^+\phi = d^-\psi$ by induction on
  $\psi$:
  \begin{align*}
    \phi \jnctx a/\alpha &= \phi\\
    \phi \jnctx (\psi,t/x,a/\alpha) &= (\phi \jnctx \psi),t/x,a/\alpha
  \end{align*}
\end{definition}

It follows by an inductive argument that this operation is associative
and unital.

\begin{lemma}[Inversion]
  \begin{enumerate}
  \item If $\Phi \vdash \psi :: (\alpha:\cat C)$ then $\Phi =
    \beta:\cat D$ for some $\cat D$ and $\psi = a/\alpha$ where
    $\beta:\cat D \vdash a : \cat C$.
  \item If $\Phi \vdash \psi :: \Psi_1 \jnctx \Psi_2$, then there
    exists unique $\Phi_1,\Phi_2,\psi_1,\psi_2$ such that $\Phi =
    \Phi_1\jnctx\Phi_2$ and $\Phi_1\vdash\psi_1 :: \Psi_1$ and
    $\Phi_2\vdash \psi_2 : \Psi_2$ and $\psi = \psi_1\jnctx\psi_2$.
  \end{enumerate}
\end{lemma}

\begin{definition}
  We define the vertical composition of transformation substitutions
  $\phi[\psi]$ inductively on $\phi$.
  \begin{align*}
    (a/\alpha)[b/\beta] &= a[b/\beta]/\alpha\\
    (\phi_1,t/x,a/\alpha)[\psi_1\jnctx\psi_2] &= \phi_1[\psi_1], t[\psi_2],a[d^+\psi_2]
  \end{align*}
  This covers all cases by \cref{inversion}.

  We define the vertical identity $\id_\Phi$ by induction on $\Phi$
  \begin{align*}
    \id_{\alpha:\cat C} &= \alpha/\alpha\\
    \id_{\Phi,x:R,\alpha:\cat C} &= \id_{\Phi},x/x,\alpha/\alpha
  \end{align*}
\end{definition}
By an inductive argument, this operation is associative and unital.

\begin{lemma}{Interchange}
  We prove the interchange law
  $$(\phi\jnctx\psi)[\phi'\jnctx\psi'] = \phi[\phi'] \jnctx \psi[\psi']$$
  by induction on $\psi$.

  \begin{enumerate}
  \item When $\psi=a/\alpha$
      \begin{align*}
    (\phi\jnctx a/\alpha)[\phi'\jnctx b/\beta]
        &= \phi[\phi']\\
        &= \phi[\phi'] \jnctx (a[b/\beta]/\alpha)\\
        &= \phi[\phi'] \jnctx (a/\alpha)[b/\beta]
  \end{align*}
    \item When $\psi=\psi,t/x,a/\alpha$
      \begin{align*}
        (\phi\jnctx (\psi,t/x,a/\alpha))[\phi' \jnctx (\psi'_1\jnctx\psi'_2)]
        &= ((\phi\jnctx \psi),t/x,a/\alpha)[(\phi' \jnctx \psi'_1)\jnctx\psi'_2]\\
        &= (\phi\jnctx \psi)[\phi' \jnctx \psi'_1],t[\psi'_2]/x,a[d^+\psi'_2]/\alpha\\
        &= (\phi[\psi'] \jnctx \psi[\psi']),t[\psi'_2]/x,a[d^+\psi'_2]/\alpha\\
        &= \phi[\psi'] \jnctx (\psi[\psi'],t[\psi'_2]/x,a[d^+\psi'_2]/\alpha)\\
      \end{align*}
  \end{enumerate}
\end{lemma}

\begin{figure}
  \begin{mathpar}
    % vertical arrows
    \inferrule*[Right=VComp*]
    {\alpha:\cat C \vdash b : \cat D \and
     \beta:\cat D \vdash c : \cat E}
    {\alpha:\cat C \vdash c[b/\beta] : \cat E}\\

    \beta[b/\beta] = b\and
    c[\beta/\beta] = c\and
    c[b/\beta][a/\alpha] = c[b[a/\alpha]/\beta]\\

    % horizontal arrows
    \inferrule*[Right=HComp*]
    {\Gamma \vdash \Phi \isavectx\\\\
      \Gamma\vdash\Psi\isavectx\\\\
      d^+\Phi = d^-\Psi}
    {\Gamma \vdash \Phi \jnctx \Psi \isavectx}

    d^+(\Phi \jnctx \Psi) = d^+\Psi\and
    d^-(\Phi \jnctx \Psi) = d^-\Phi\\

    \Phi \jnctx (d^+\Phi) = \Phi \and
    (d^-\Phi) \jnctx \Phi = \Phi \and
    (\Phi \jnctx \Psi) \jnctx \Sigma = \Phi \jnctx (\Psi \jnctx \Sigma)\\

    % 2-cells
    %% vertical composition
    \inferrule*[Right=V2Comp*]
    {\Sigma \vdash \psi :: \Psi \and
     \Psi \vdash \phi :: \Phi}
    {\Sigma \vdash \phi[\psi] :: \Phi}

    \inferrule*[Right=V2Id*]
    {}
    {\Phi \vdash \id_\Phi :: \Phi}\\

    \id_\Phi[\phi] = \phi\and
    \phi[\id_\Psi] = \phi\and
    \phi[\psi][\sigma] = \phi[\psi[\sigma]]\\

    %% Horizontal composition
    \inferrule*[Right=H2Comp*]
    {\Phi' \vdash \phi :: \Phi \\\\
     \Psi' \vdash \psi :: \Psi \\\\
     d^+\phi = d^-\psi 
    }
    {\Phi' \jnctx \Psi' \vdash \phi \jnctx \psi :: \Phi \jnctx \Psi}

    d^-(\phi\jnctx\psi) = d^-\phi\and
    d^+(\phi\jnctx\psi) = d^+\phi\and

    \phi\jnctx d^+\phi = \phi\and
    d^-\phi \jnctx \phi = \phi\and
    \phi\jnctx(\psi\jnctx\sigma) = (\phi\jnctx\psi)\jnctx\sigma

    %% Interchange
    \inferrule*[Right=Interchange]
    {\Phi'' \vdash \phi' :: \Phi' \and
     \Psi'' \vdash \psi' :: \Psi' \and
     d^+\phi' = d^-\psi'\\\\
     \Phi' \vdash \phi :: \Phi \and
     \Psi' \vdash \psi :: \Psi \and
     d^+\phi = d^-\psi
    }
    {(\phi\jnctx\psi)[\phi'\jnctx\psi'] = \phi[\phi'] \jnctx \psi[\psi']}\\

    % Interaction with displays
    \inferrule
    {\Psi \vdash \phi :: \Phi\and
     \Phi \vdash t : R[a/\alpha;b/\beta]}
    {\Psi\vdash t[\phi] : R[a[d^-\phi]/\alpha;b[d^+\phi]/\beta]}

    t[\id_\Phi] = t\and
    t[\phi][\psi] = t[\phi[\psi]]
  \end{mathpar}
  \caption{Syntactic Double Category with Displays Structure}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\cdot : \Gamma \to \cdot}

    \inferrule
    {\gamma : \Delta \to \Gamma \and \Delta \vdash M : A[\gamma]}
    {\gamma,M/X : \Delta \to \Gamma,X:A}

    \inferrule
    {\Gamma \vdash \Phi \isavectx \and
    \Gamma \vdash \Psi \isavectx \and
    d^+\Phi = d^- \Psi}
    {\Gamma \vdash \Phi \jnctx \Psi \isavectx}

    \inferrule
    {\Gamma \vdash \phi : \Phi' \to \Phi \and
     \Gamma \vdash \psi : \Psi' \to \Psi\and
     d^+\phi = d^- \psi}
    {\Gamma \vdash \phi \jnctx \psi : \Phi' \jnctx \Psi' \to \Phi \jnctx \Psi}

    \inferrule*[right=ObSubst]
    {\alpha : \cat C \vdash B : \cat D\and
    \beta : \cat D \vdash A : \cat E}
    {\alpha : \cat C \vdash A[B/\beta] : \cat E}\\

    \inferrule*[right=Everything Subst]{\gamma : \Delta \to \Gamma \and \Gamma J}{\Delta J}

    \inferrule{\Gamma \vdash \xi : Z \to \Xi\and \Gamma \pipe \Xi \vdash R \isaSet}{\Gamma \pipe Z \vdash R[\xi] \isaSet}

    \inferrule{\Gamma \vdash \phi : \Psi \to \Phi\and \Gamma \pipe \Phi \vdash M : R}{\Gamma \pipe \Psi \vdash M[\phi] : R[\underline \phi]}
  \end{mathpar}
  \begin{align*}
    (\Phi \jnctx \Psi) \jnctx \Sigma &= \Phi \jnctx (\Psi \jnctx \Sigma)\\
    \Phi \jnctx \alpha:C &= \Phi\\
    \alpha:C \jnctx \Psi &= \Psi\\\\
    d^\pm(A/\alpha) &= A/\alpha\\
    d^+(\phi_1,M/x,\phi_2) &= d^+ \phi_2\\
    d^-(\phi_1,M/x,\phi_2) &= d^- \phi_1\\\\
    \underline \cdot &= \cdot\\
    \underline {A/\alpha} &= A/\alpha\\
    \underline {\phi_1,M/x,\phi_2} &= d^-\phi_1;d^+\phi_2
  \end{align*}
  \caption{Admissible Rules}
\end{figure}
